---
title: "OrthoPaiR Package Vignette"
author: "Tomoyuki Furuta"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{OrthoPaiR Package Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  ---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

# Introduction

`OrthoPaiR` is an R package for identifying **syntenic orthologous genes** across multiple genomes.
The main user-facing workflow consists of:

1. Preparing an input table for each genome.
2. Reformatting inputs with `reformatFiles()`.
3. Running reciprocal search with `rbh()`.
4. Running ortholog inference with `orthopair()`.
5. Optionally summarising all pairs with `reorgOrthopairs()`.

This vignette explains how to prepare inputs, run the pipeline for two or more genomes, and interpret the output files.

# Prerequisites

```{r, eval=FALSE}
install.packages("devtools")
install.packages("BiocManager")

BiocManager::install(c(
  "GenomicRanges", "GenomeInfoDb", "GenomicFeatures",
  "BiocGenerics", "Biobase", "IRanges",
  "rtracklayer", "Biostrings", "S4Vectors", "BSgenome"
))

install.packages(c("parallel", "ggplot2", "dplyr"))
```

External command‑line tools:

- **BLAST+** (`makeblastdb`, `blastn` in `blast_path`) – required for DNA sequence similarity; `blastp` is required only when `use_prot = TRUE` for protein sequence similarity.
- **DIAMOND** (`diamond` in `diamond_path`) – required when `use_prot = TRUE` for protein sequence similarity.
- **Miniprot** (optional; only if `run_miniprot = TRUE` to predict missing ORFs) – requires genome FASTA files in `orgInputFiles()`.

# Preparing input files

Create a data.frame with one row per genome and columns:
`name`, `genome`, `gff`, `cds`, `prot`.
Then pass it to `reformatFiles()`.

```{r, eval=FALSE}
library(OrthoPaiR)

input_df <- data.frame(
  name = c("Osat", "Hvul"),
  genome = c("path/to/osat_genome.fa", "path/to/hvul_genome.fa"),
  gff = c("path/to/osat.gff3", "path/to/hvul.gff3"),
  cds = c("path/to/osat_cds.fa", "path/to/hvul_cds.fa"),
  prot = c(NA, NA),
  stringsAsFactors = FALSE
)
```

If you prefer incremental construction, you can use `orgInputFiles()`:

```{r, eval=FALSE}
in_list <- orgInputFiles(
  name = "Osat",
  genome = "path/to/osat_genome.fa",
  gff = "path/to/osat.gff3",
  cds = "path/to/osat_cds.fa",
  prot = NA
)

in_list <- orgInputFiles(
  object = in_list,
  name = "Hvul",
  genome = "path/to/hvul_genome.fa",
  gff = "path/to/hvul.gff3",
  cds = "path/to/hvul_cds.fa",
  prot = NA
)

input_df <- as.data.frame(in_list, stringsAsFactors = FALSE)
```

## Required GFF and FASTA formats

- **GFF3 file**
  - Must contain at least `gene` and transcript features (`mRNA` or `transcript`) and `CDS` features.
  - Must have a non‑empty `gene_id` column with no `NA`.
  - The `Parent` column of transcripts must match the `ID` of genes one‑to‑one (and vice versa).
- **CDS FASTA**
  - FASTA record names must correspond to transcript IDs used in the GFF.
  - All CDS names must be present in the GFF; otherwise `reformatFiles()` stops with an error.
- **Protein FASTA** (only if `prot` is provided)
  - Protein record names must also match `ID` values in the same GFF.
- **Genome FASTA**
  - Required when CDS sequences are reconstructed from GFF.
  - If supplied, all `seqlevels(gff)` must be present in the genome FASTA sequence names.

## Reformatting input files with `fixInputFiles()`

If your input files do not meet the requirements above, you can use `fixInputFiles()` to reformat them. This function helps prepare GFF, CDS, protein, and genome FASTA files to be compatible with OrthoPaiR.

```{r, eval=FALSE}
fixed_files <- fixInputFiles(
  genome = "path/to/genome.fa",    # optional
  gff    = "path/to/annotation.gff",
  cds    = "path/to/cds.fa",       # optional
  prot   = "path/to/prot.fa",      # optional
  autofix = FALSE                  # TRUE for automatic fixes
)
```

The function performs the following operations:

- **Genome FASTA reformatting** (if provided):
  - Renames sequence names to match GFF seqlevels (with interactive prompts or automatic renaming if `autofix = TRUE`).
  - Allows padding chromosome numbers and adding prefixes (e.g., "chr01", "chr02").
  - Enables removal of unwanted sequences.

- **GFF file reformatting**:
  - Drops unnecessary feature types, keeping only `gene`, `mRNA`/`transcript`, `exon`, and `CDS` features.
  - Ensures proper `ID` and `Parent` relationships between features.
  - Sets the `gene_id` column for all features.
  - Fixes `Parent` column entries to contain single values (required by OrthoPaiR).
  - Updates sequence names (seqlevels) to match the reformatted genome FASTA.

- **CDS and protein FASTA reformatting** (if provided):
  - Attempts to match FASTA record names with GFF `ID` values using multiple strategies:
    - Direct matching with `ID`, `transcript_id`, or `protein_id` columns.
    - Pattern-based matching (removing suffixes, extracting IDs from complex names).
  - If `autofix = TRUE`, automatically tries multiple matching strategies.
  - If matching fails and a genome FASTA is provided, creates new CDS and protein FASTA files from the GFF and genome sequences.

The function returns a list with paths to the reformatted files (with `.op_fixed` suffix):
- `fixed_files$gff`: reformatted GFF file
- `fixed_files$cds`: reformatted CDS FASTA (if provided)
- `fixed_files$prot`: reformatted protein FASTA (if provided)
- `fixed_files$genome`: reformatted genome FASTA (if provided)

You can then use these fixed files in the `input_df` passed to `reformatFiles()`:

```{r, eval=FALSE}
input_df <- data.frame(
  name = "Osat",
  genome = fixed_files$genome,
  gff = fixed_files$gff,
  cds = fixed_files$cds,
  prot = fixed_files$prot,
  stringsAsFactors = FALSE
)
```

**Note**: When `autofix = FALSE` (default), the function will prompt you interactively for decisions about renaming sequences and fixing IDs. Set `autofix = TRUE` for automatic fixes, but be aware that automatic fixes may not always succeed if the input format is too different from expected formats.

# Running the workflow

Run the current workflow as four explicit steps:

```{r, eval=FALSE}
wd <- "path/to/output/orthopair"

reformatFiles(
  object = input_df,
  working_dir = wd,
  overwrite = TRUE,
  n_threads = 8
)

rbh(
  working_dir = wd,
  blast_path = "/path/to/blast/bin",
  target_pair = NULL,
  n_threads = 8,
  overwrite = TRUE
)

orthopair(
  working_dir = wd,
  n_threads = 8
)

reorgOrthopairs(
  working_dir = wd,
  rename = TRUE,
  n_threads = 8
)
```

- `reformatFiles()` writes per-genome normalized files under `working_dir/input`.
- `rbh()` writes reciprocal hit files under `working_dir/rbh`.
- `orthopair()` writes pairwise ortholog TSV files under `working_dir/orthopair`.
- `reorgOrthopairs()` writes graph and orthogroup outputs under `working_dir/reorg_out`.

## Accessing pairwise outputs

After running `orthopair()`, use helper accessors:

```{r, eval=FALSE}
# Genome/sample ID mapping from input folders
getGenomeID(working_dir = wd)

# Pairwise ortholog table
op <- getOrthoPair(
  working_dir = wd,
  pair = c(1001, 1002),
  score = TRUE,
  loc = TRUE
)

# Pairwise orphan genes
orphan <- getOrphan(
  working_dir = wd,
  pair = "1001_1002"
)

# Summary table for one pair
summaryOrthoPair(
  working_dir = wd,
  pair = c(1001, 1002)
)
```

## Two‑genome vs multi‑genome use

When you analyse **only two genomes**, `orthopair()` output in `working_dir/orthopair` is usually sufficient.
For **three or more genomes**, run `reorgOrthopairs()` to generate graph-level summaries.

# Example: four‑genome analysis (sample script)

```{r, eval=FALSE}
root_dir   <- "/path/to/workspace/orthology"
input_dir  <- file.path(root_dir, "input")
output_dir <- file.path(root_dir, "output")

library(OrthoPaiR)

in_list <- orgInputFiles(
  name = "Osat",
  genome = NA,
  gff = file.path(input_dir, "osat.gff"),
  cds = file.path(input_dir, "osat_cds.fa"),
  prot = file.path(input_dir, "osat_prot.fa")
)

in_list <- orgInputFiles(
  object = in_list,
  name = "Hvul",
  genome = NA,
  gff = file.path(input_dir, "hvul.gff"),
  cds = file.path(input_dir, "hvul_cds.fa"),
  prot = file.path(input_dir, "hvul_prot.fa")
)

in_list <- orgInputFiles(
  object = in_list,
  name = "Atha",
  genome = NA,
  gff = file.path(input_dir, "atha.gff"),
  cds = file.path(input_dir, "atha_cds.fa"),
  prot = file.path(input_dir, "atha_prot.fa")
)

in_list <- orgInputFiles(
  object = in_list,
  name = "Mpol",
  genome = NA,
  gff = file.path(input_dir, "mpol.gff"),
  cds = file.path(input_dir, "mpol_cds.fa"),
  prot = file.path(input_dir, "mpol_prot.fa")
)

wd <- file.path(output_dir, "orthopair")

input_df <- as.data.frame(in_list, stringsAsFactors = FALSE)
reformatFiles(object = input_df, working_dir = wd, overwrite = TRUE, n_threads = 30)
rbh(working_dir = wd, blast_path = "/path/to/blast/bin", n_threads = 30, overwrite = TRUE)
orthopair(working_dir = wd, n_threads = 30)
reorgOrthopairs(working_dir = wd, rename = TRUE, n_threads = 30)
```

This script corresponds to the contents of `sample_data/orthopair`, which you can use as a reference for expected outputs.

# Output files produced by `orthopair()`

The sample output in `sample_data/orthopair` illustrates the directory structure created by the pipeline:

When you run the pipeline, the output directory will have the following structure (example for four genomes and default parameters):

```
/your_working_dir/
├── blast/
│   ├── *_group_cds.fa
│   ├── *.blastdb*
│   ├── *_blast.out
│   └── ...
├── rbh/
│   ├── all.rbh.tsv
│   ├── 1001_1002.rbh
│   └── ...
├── orthopair/
│   ├── 1001_1002.tsv
│   ├── 1001_1003.tsv
│   ├── orthopair_pairwise_mutual_ci_stats.tsv
│   └── orthopair_genome_mean_mutual_ci_matrix.tsv
├── reorg_out/
│   ├── orthopair.graphml
│   ├── orthopair_list.tsv
│   └── pairwise/
│       ├── 1001_1002.tsv
│       └── ...
├── input/
│   ├── 1_Osat/
│   ├── 2_Hvul/
│   └── ...
└── ... (other possible intermediate or log files)
```

Detailed descriptions of the output files:

- **`input/<index>_<name>/`**
  - `gff.rds`, `gff_df.rds`, `cds.fa` and related normalized inputs.
- **`rbh/`**
  - `all.rbh.tsv` and split pair files (`<genomeA>_<genomeB>.rbh`).
- **`orthopair/`**
  - Pairwise ortholog tables (`<genomeA>_<genomeB>.tsv`) and pairwise mutual CI summaries.
  - Pairwise orphan tables in `orthopair/orphan/<genomeA>_<genomeB>.tsv`.
- **`reorg_out/`**
  - `orthopair.graphml`, `orthopair_list.tsv`, and pairwise tables in `reorg_out/pairwise/`.
- **`reorg_orphan/`**
  - Pairwise orphan tables computed from `reorg_out/pairwise`.

